Response time of a plasmonic distributed feedback laser in large-signal modulation regime

### 2019-04-04 Response_DFB_Pulse_PRApplied.md

**Review**

*However, the manuscript focused on plasmonic nanohole arrays with a specific parameter design (100-nm hole, 470-nm spacing and 100-nm film thickness). Are the conclusions on the critical pump spot size and ultrafast response limit (1 ps) general enough to plasmon nanohole array lasers in different design parameters (film thickness, hole size, spacing)? Also, plasmon lasers based on nanohole arrays and nanoparticle arrays show different lasing mechanisms from surface plasmon polariton and surface lattice resonances, respectively. It is not a fair comparison between the calculated response time of plasmon nanohole arrays and experimental results of nanoparticle arrays to verify the accuracy of theoretical modeling (final paragraph, the introduction section).*

**Response**

Note that the pump spot effectively excites the EM field inside the pumped area which forms an effective laser mode. The existence of the *critical* size of the pump beam corresponds to the minimal size of the effective mode. Qualitatively, the minimal size of the effective mode is equal to the characteristic propagation length of the EM field along the plasmonic structure, which is$\sim v_{\text{group}}/ \gamma_a$. When the size of the pump beam is greater than the size of the effective laser mode, the response time almost does not depend on the size of the pump beam. When the size of the pump beam becomes smaller than the size of the effective laser mode, the response time increases with the decrease of the pump beam size.

The effective laser mode is formed by the Bloch modes of a periodic plasmonic structure which determine a characteristic group velocity and a dissipation constant. Of course, modes in different types of structure have different group velocities and relaxation constants. Nevertheless, based on the qualitative arguments presented above, we can conclude that there is always the *critical* size of the pump beam, for which the response time is minimal.

However, the specific value of the time response strongly depends on the type of the plasmonic structure. Indeed, the response time depends on the Rabi constant of interaction. In turn, the Rabi constant is determined by the volume of the plasmonic mode which is different for surface plasmon-polariton and surface lattice resonances. In our numerical simulations, we consider plasmon lasers based on nanohole arrays, though in the work [[Daskalakis, Konstantinos S., et al. *Nano letters* 18.4 (2018): 2658-2665.](https://pubs.acs.org/doi/abs/10.1021/acs.nanolett.8b00531)] the experiment has been conducted for plasmon lasers based on nanoparticle arrays. In this regard, we agree with the referee that the comparison of the response times obtained in the experiment and in our simulations is not a fair verification of the accuracy of our theoretical modeling.

In the revised version of the manuscript, we reformulate the last paragraph in the <u>Introduction</u> section by excluding this direct comparison between our theory and the experiment.

---

@@

**Response**

In our modeling, we take into account a large number of atoms, which have different coordinates $\textbf{r}_m​$. The coupling constant between $j​$th mode of the plasmonic structure and $m​$th atom depends on atom's coordinates as
$$
\Omega _{jm} =  - {{\bf{d}}_m} \cdot {{\bf{E}}_j}({{\bf{r}}_m})/ \hbar
\tag{1}
$$

where the electric field is taken at the position of the $m$th atom. The Rabi constants enter as coefficients in the Eqs. (1) – (3).

The pulse pump of the active medium enters as an initial condition for the atomic population inversion. Because interaction of each atom with plasmon EM field depends on the atomic coordinate, a distribution of the initial population inversion over the atoms allows to determine the pump size. The initial value of the population inversion for the atoms whose coordinates lie inside the area of the pumping spot is $D_0​$. Initial population inversions of other atoms equal to $-1​$, i.e. they are in the ground state. The *size* of the pump beam enters as an area, inside which atoms are excited.

Pump power density (excitation fluence) enters through the value of $D_0​$. The connection between them has the form 
$$
\frac{D_0 + 1}{2} = \eta \frac{P_0 \Delta t}{n_c S l \hbar \omega}
\tag{2}
$$
Here $P_0$ is power of the laser pulse pump; $\Delta t$ is duration of the pump pulse; $n_c$ is the concentration of atoms of the active medium; $S$ is the area of the pump beam; $l$ is the width of the active medium layer; $\hbar \omega$ is the transition frequency of the atom; and $(D_0 + 1)/2$ is the population of the excited state arising due to the pump pulse. In real experiment, there are additional circumstances as a result of which a part of the energy from pulse pumping does not come into the active medium. These circumstances are taken into account as an additional coefficient $\eta <1$ in a right-hand side of (2).  In our simulations, we assume that $P_0$ is such that all atoms inside the pumped area are inverted, namely we take the values from $D(0) = 0.25$ to $D(0) = 1$.

@@

*Did the authors consider any diffractive coupling between nanoholes in the equations of system dynamics, given that pump size may influence the resonance mode quality (Purcell factor) and EM fields near the holes?*

**Response**

Diffractive coupling between nanoholes is taken into account in laser equations (1) – (3).

First, we have found the EM field distribution of the Bloch modes of the plasmonic structure within the coupled-mode theory. Note that the coupled-mode theory describes the hybridization of initially non-interacting SPP modes through scattering on the holes, i.e. takes into account diffractive coupling [[V. T. Tenner, A. N. van Delft, M. J. A. de Dood, and M. P. van Exter, *Journal of Optics* 16, 114019 (2014)](https://iopscience.iop.org/article/10.1088/2040-8978/16/11/114019/meta)]. Scattering rates have been taken from Ref. [[V. T. Tenner, A. N. van Delft, M. J. A. de Dood, and M. P. van Exter, *Journal of Optics* 16, 114019 (2014)](https://iopscience.iop.org/article/10.1088/2040-8978/16/11/114019/meta)] for the plasmonic laser based on Au film perforated by the nanoholes array. The coupled-mode theory with such scattering rates demonstrates good agreement with the experimental measurement of the band-structure of the plasmonic nanohole arrays laser [[V. T. Tenner, A. N. van Delft, M. J. A. de Dood, and M. P. van Exter, *Journal of Optics* 16, 114019 (2014)](https://iopscience.iop.org/article/10.1088/2040-8978/16/11/114019/meta)].

Second, we use the obtained EM field distributions to calculate the Rabi constants between the Bloch modes and atoms of the active medium. These Rabi constants, the eigenfrequencies and the relaxation rates of the Bloch modes enter the laser equations (1) – (3). The Eqs. (1) – (3) determine the amplitude of each Bloch mode. The amplitudes depend on the interaction between the Bloch mode and active atoms, and, as a consequence, on the pump beam size. The field of the collective mode is formed by the large number of Bloch modes of a periodic plasmonic structure. As a result, the volume of the collective mode changes with a change in the pump beam size. The Purcell factor and the rates of the spontaneous and stimulated emissions also change.

@@

**Review**

*Also, how is the Rabi constant in the modeling related to the Purcell factor of a specific plasmon cavity? More justifications are needed.*

**Response**

We added the derivation of the Rabi constant through the parameters of the plasmonic structure and the active medium in the revised version of the manuscript , see <u>Section 2</u> in main part of the text and <u>Appendix</u>.

---

@@

*We have other comments below*

**1.** *On page 4, the authors claimed that “The time independence of the EM field distribution allows us to consider the plasmonic laser as a single mode laser,” but they need to justify how the time evolution of EM field is correlated to single-mode approximation. Is there any effect from high-order diffractive modes for the hole arrays? Why does the EM field distribution slightly change from Fig 3d to 3f?*

**Response**

In our simulations we take into account $\sim 500$ Bloch modes of the plasmonic structure which form together the EM field distribution in the active medium. However, our simulations show that the *resulting* EM field distribution slightly changes in the first and second stage of the output pulse evolution. In other words, there is a dynamical equilibrium between relaxation and propagation of the EM field outside the pump beam, on the one hand, and the stimulated amplification of the EM field inside the pump beam, on the other hand. We use the fact of the constancy of the EM field distribution to construct the single-mode model. We consider that there is a single collective mode, the EM field distribution of which coincides with the EM field distribution found in the numerical simulations of the multimode problem. This assumption allows to derive the single-mode laser equations. The results obtained in the framework of the single-mode model are in good agreement with the results of the multimode model, see Figs. 2(a), 4 and Fig. 5a.

@@

**Review**

*My main criticism of the paper concerns the statement that the calculations reveal an "optimal" beam size, yielding the shortest response time. From Fig.2(b) one sees that the shortest response time corresponds to a very flat minimum in the red curve. I doubt the robustness of this minimum as the system's parameters, or the level of approximation of the theory, are changed. I strongly suggest that the Authors rephrase the text, claiming that $d \sim 15 \mu\text{m}​$ corresponds to the minimum size to achieve a short response time, and not the optimal size.*

**Response**

We agree with the referee that the minimum is flat and some unaccounted effects, such as non-ideal plasmonic structure, inhomogeneous distribution of the active medium, etc., may change the value of the size beam for which the response time is minimal. We rephrase the text claiming that $d \sim 15 \mu \text{m}$ corresponds to the minimum size to achieve the short response time for considered structure, rather than the optimal size.

---

### 2019-04-04 DFB_Pulse_LaTeX/DFB_pulse_inProcess.tex
Final version of Response

	The time response to an external signal is the main characteristic of optoelectronic devices which determines their maximum modulation frequency.
	The use of plasmonic structures gives the opportunity to significantly reduce the response time.
	In this paper, we study the temporal dynamics of a plasmonic distributed feedback laser consisting of a two-dimensional metallic structure coated with a layer of the active medium.
	Combining numerical simulations with an analytical evaluation we show that the response time in the large-signal modulation regime strongly depends on the size of the pump beam.
	We demonstrate that the interaction of a large number of modes of the plasmonic structure with the pumped active medium leads to the formation of a single collective mode.
	If the size of the pump beam is larger than the propagation length of electromagnetic waves in the plasmonic structure, the collective mode is localized in the pumped area.
	In this case, the response time slightly increases with the size of the pump beam.
	In the opposite case, the mode is not localized inside the pumped area and the response time grows with the narrowing of the beam.
	We demonstrate that there is a critical minimum size of the pump beam for which the modulation speed is a maximum and can achieve $1$ THz.
	The obtained result opens the way to increasing both the modulation speed and the energy efficiency of plasmonic laser devices.

@@

Here ${{\bf{d}}_m}$ is the dipole moment of $m$th atom at the transition frequency and ${{\bf{E}}_j}({{\bf{r}}_m})$  is the electric field `per one quantum' in $j$th mode at the position of $m$th atom. Every atom of the active medium has its coordinates, which determine the Rabi constants of interaction between the atom and the Bloch modes of the plasmonic structure, see Eq.~(\ref{eq4SM}). This allows us to take into account the spatial distribution of the pumping and the population inversion within the active medium. Assuming that the time of the external pump pulse lies in the femtosecond range and is much shorter than all the characteristic times of the plasmonic DFB laser, in our model, the ultrashort pulse pumping is described as the initial condition, $D_m(0)$ for atoms lying inside the pump beam. The value of $D_m(0)$ is proportional to the pumping energy per unit area. The population inversions of the atoms whose coordinates lie outside the pump beam are $-1$ at the initial time. Note that the square of the Rabi constant is inversely proportional of the Bloch mode volume and is proportional of the Purcell factor of the mode (see Appendix).

@@

For a quantitative description of the temporal dynamics of a plasmonic DFB laser under pulse pumping, we consider a laser based on an Au film perforated by a periodic array of nanosized holes and coated with the layer of the active medium, see Fig.~\ref{fig1}. We consider that the thickness of the Au film is $100$ $\text{nm}$, the diameter of the holes is $160$ $\text{nm}$, and the distance between holes is $470$ $\text{nm}$. The surface of the entire periodic plasmonic structure is $190 \times 190$ $\mu \text{m}^2$. We consider that the transition frequency of the active medium, $\omega_{\sigma}$, is close to the frequency of the second Bloch condition in the plasmonic structure, i.e. $k_\text{pl}(\omega_{\sigma}) \approx G$, where $k_\text{pl}$ is a wavevector of the mode in the plasmonic structure and $G$ is a reciprocal lattice vector \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot}, and the linewidth of the active medium is about $100 \text{nm}$.

A similar plasmonic DFB laser have been implemented in a number of experiments \cite{BeijnumPRL,TennerJOpt,TennerACSPhot} and its parameters have been well studied \cite{BeijnumPRL,TennerJOpt,TennerACSPhot,nefedkin2018acsphot}.

The EM field distribution of the Bloch modes of the plasmonic structure have been found within the coupled-mode theory \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot} (see also Appendix), which describes the hybridization of modes of the system without holes through scattering on the holes. Scattering rates for these modes have been taken from \cite{TennerJOpt,TennerACSPhot}. The coupled-mode theory with such scattering rates demonstrates a good agreement with the experimental measurement of the band-structure of the plasmonic nanohole array laser \cite{TennerJOpt,TennerACSPhot}.

In numerical simulations, we take into account $468$ Bloch modes of the plasmonic structure lying on different dispersion curves. The step of quantization of the Bloch wavevectors is determined by the size of the periodic plasmonic structure \cite{nefedkin2018acsphot} and Suppl.Mat. of \cite{nefedkin2018acsphot}.

@@

We use the fact of the constancy of the EM field distribution to construct a single-mode model. To do this, we assume that there is a single collective mode, the EM field distribution of which coincides with the EM field distribution found in the numerical simulations of the multi-mode problem. Furthermore, we assume that the pumped active medium interacts only with such a single collective mode. Within these approximations, the system dynamics can be approximately described by the following system of equations \cite{SiegmanLasers}:
\begin{gather} 
\dot n =  - 2{\gamma _a}n + 2{N_{\text{at}}}{\Omega _R}\varphi \label{ZE1}
\\
\dot \varphi = - \left( {{\gamma _a} + {\gamma _\sigma }} \right)\varphi + {\Omega _R}\left( {2nD + D + 1} \right)/2 \label{ZE2}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 4{\Omega _R}\varphi, \label{ZE3}
\end{gather}
where $n$ is the number of photons in the collective mode, $D$ is the mean population inversion of pumped atoms of the active medium, $\varphi$ describes the mean energy flow from the active medium to the collective mode, and $N_{\text{at}}$ is the number of atoms of the active medium lying in the collective mode.
We suppose that the relaxation rate of the EM field in the collective mode is equal to the characteristic relaxation rate of the modes of the plasmonic structure, i.e., $\gamma_a \simeq 10^{12} \text{s}^{-1}$.
The typical relaxation rate of the population inversion of the active medium is $\gamma_D \simeq 10^9 \text{s}^{-1}$ while the phase relaxation rate is $\gamma_{\sigma}\simeq 10^{13} \text{s}^{-1}$.
Thus, the phase relaxation rate is much larger than other relaxation rates, $\gamma_\sigma \gg \gamma_a, \gamma_D$.
For this reason, one can adiabatically eliminate $\varphi$ from Eqs.~(\ref{ZE1})--(\ref{ZE3}) and obtain the conventional rate equations
\begin{gather} 
\dot n =  - 2 {\gamma _a}n + {N_{\text{at}}}\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE1}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 2\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE2}
\end{gather}
As initial condition we take $D(0) = 1$ and $n(0) = 0$, i.e., all atoms are inverted and there are no photons in the mode.
The temporal dynamics of such a system is well-known \cite{SiegmanLasers}; it is shown in Fig.~\ref{fig4}.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\linewidth}
		\includegraphics[width=\linewidth]{Fig5a.eps}
		%\caption{}
	\end{subfigure}
	\begin{subfigure}[h]{0.45\linewidth}
		\includegraphics[width=\linewidth]{Fig5b.eps}
		%\caption{}
	\end{subfigure}
	\caption{(a) The dependence of the diameter of the collective mode on the pump beam diameter (blue line). The black dashed line denotes the pump beam diameter. (b) The dependence of the delay time on the pump beam diameter. The blue solid line is obtained from numerical simulation of Eqs. (\ref{eq1})--(\ref{eq3}), the orange thick line is obtained from Eq. (\ref{tdel}).
	}
	\label{fig5}
\end{figure}

The figure \ref{fig5}b shows that the analytical approximation, Eq.~(\ref{tdel}), agrees well with the numerical simulation of the exact model, Eqs. (\ref{eq1})--(\ref{eq3}).
Let us analyze the dependence of the delay time, Eq.~(\ref{tdel}), on the size of the pumped area.
The number of atoms depends quadratically on the diameter $d_{\text{beam}}$ of the pump beam, i.e., $N_{\text{at}} \sim d_{\text{beam}}^2$.
In turn, the square of the Rabi constant is inversely proportional to the square of the diameter $d_{\text{mode}}^2$ of the collective mode, i.e., $\Omega_{R}^{2} \sim 1/d_{\text{mode}}^2$.
So, we have $t_{\text{delay}} \sim d_{\text{mode}}^2\ln \left( d_{\text{beam}}^2 \right ) /d_{\text{beam}}^2$.
The dependence of $d_{\text{mode}}$ on $d_{\text{beam}}$ is shown in Fig.~\ref{fig5}a.

It can be seen that there is a critical value, $d_{\text{beam}}^*$, of the diameter of the pump beam, which divides two different cases.
When $d_{\text{beam}}>d_{\text{beam}}^*$, the electric field of the collective mode is localized inside the pump beam, see Figs.~\ref{fig3}(d)--(f).
As a consequence, the mode diameter is proportional to the beam diameter, i.e., $d_{\text{mode}} \sim d_{\text{beam}}$, cf. Fig.~\ref{fig5}a.
For this case, from Eq.~(\ref{tdel}) we have $t_{\text{delay}} \sim \ln d_{\text{beam}}$.
This dependence is in agreement with a slow growth of the delay time of the DFB laser, see Fig\ref{fig2}b.
When $d_{\text{beam}}<d_{\text{beam}}^*$, the electric field of the collective mode is localized outside the pump beam, see Figs.~\ref{fig3}(a)--(c), and the mode size is determined by the characteristic propagation length of the EM field in the unpumped region.
As a result, the mode diameter does not depend on the beam diameter, and $d_{\text{mode}} \sim \text{const}$, see Fig.~\ref{fig5}a.
In this case, $t_{\text{delay}} \sim \ln(d_{\text{beam}}^2)/d_{\text{beam}}^2$.
This is in agreement with an increase of the delay time  when the pump beam diameter of the plasmonic DFB laser is reduced, see Fig.~\ref{fig2}b.
Note that the critical value $d_{\text{beam}}^*$ corresponds with the characteristic propagation length of the EM field outside the pumped active medium and may be evaluated as $d_{\text{beam}}^* \sim v_{\text{group}}/\gamma_a$.
For the parameters of the DFB laser we have $d_{\text{beam}}^* \sim 10 \mu \text{m}$, which is in agreement with the full model, Fig.~\ref{fig2}b.
Thus, because the delay time grows when $d_{\text{beam}}>d_{\text{beam}}^*$ and grows in the opposite case as well, we come to the conclusion that there is a critical value, $\sim d_{\text{beam}}^*$, for the diameter of the pump beam for which the response time is a minimum.

@@

To summarize, we have studied the temporal dynamics of a two-dimensional plasmonic DFB laser under ultrashort pulse pumping.
Relying on numerical simulations and analytical calculations, we have shown that the response time of such a laser for typical experimental parameters can be decreased to one picosecond, which corresponds to a modulation frequency of $1$ THz.
This value is close to the upper limit for plasmonic lasers (the total response time can not be shorter than the time of the third stage, which is determined by the losses in the cavity, and is $ \sim 1$ ps).

It has been shown that the laser's response time can be shortened by optimizing the size of the pump beam, as a result of the consequent reduction of the delay time.
The dependence of the response time on the size of the pump beam is for following reason.
The interaction of the laser modes with the pumped active medium results in the emergence of a single collective mode.
If the size of the pump beam exceeds the propagation length of the EM waves, this collective mode is located in the pumped area.
The response time then grows logarithmically with increasing pump beam diameter.
In contrast, when the mode is outside the pumped region, the response time is approximately inversely proportional to the square of the pump beam diameter, and so increases rapidly as this diameter decreases from its critical value.
It is interesting to note that the delay time, Eq.~(\ref{tdel}), resembles the dependence of the delay time of a superradiant burst \cite{gross1982superradiance, andreev1980collective,nefedkin2017badcavitySR}.
Indeed, it is well known \cite{gross1982superradiance, andreev1980collective} that a system consisting of initially inverted atoms placed in a low-Q cavity, i.e., when $\gamma_a \gg \gamma_D$, exhibits a superradiant burst with the delay time, Eq.~(\ref{tdel}). In the plasmonic cavity, the inequality $\gamma_a \gg \gamma_D$ is always hold true and, for this reason, the superradiance problem and the problem that we investigate here have the same physical grounds. We believe that further study of the relationship between these problems may be important in the context of creating ultrafast optoelectronic devices.

Thus, there is the diameter of the pump beam, $d_{\text{beam}}^*$, necessary to achieve the short response time. This diameter is about the decay length of the EM field in the plasmonic structure and for the considered plasmonic laser is $\sim 15 \mu$m. The existence of the critical pump beam size gives the opportunity to reduce the energy consumption because further increase of the pump beam size does not shorten the response time.
In turn, if the area of the pump beam is small, there is the opportunity to reduce the size of the device.
Thus, the obtained results pave the way to creating ultrafast optoelectronic devices based on plasmonic lasers with low power consumption which have the potential of on-chip integration.

@@

\section*{Appendix A}
To normalize the electric field ${{\bf{E}}_j}({{\bf{r}}_m})$, we equate the energy of one quantum to the energy of the electric field in the resonator:
\begin{equation}
\hbar {\omega _j} = \frac{1}{{8\pi }}\int\limits_V {\left( {{{\left. {\frac{{\partial {\mathop{\rm Re}\nolimits} \left( {\varepsilon(\omega, \textbf{r}) \omega } \right)}}{{\partial \omega }}} \right|}_{\omega  = {\omega _j}}}{{\left| {{{\bf{E}}_j}\left( {\bf{r}} \right)} \right|}^2} + {{\left| {{{\bf{H}}_j}\left( {\bf{r}} \right)} \right|}^2}} \right){d^3}{\bf{r}}} \tag{A.1}
\end{equation}
where $\varepsilon (\omega, \textbf{r})$  is the dielectric constant of the active plasmonic structure.

>>>>

	The time response to an external signal is the main characteristic of optoelectronic devices which determines their maximum modulation frequency.
	The use of plasmonic structures gives the opportunity to significantly reduce the response time.
	In this paper, we study the temporal dynamics of a plasmonic distributed feedback laser consisting of a two-dimensional metallic structure coated with a layer of the active medium.
	Combining numerical simulations with an analytical evaluation we show that the response time in the large-signal modulation regime strongly depends on the size of the pump beam.
	We demonstrate that the interaction of a large number of modes of the plasmonic structure with the pumped active medium leads to the formation of a single collective mode.
	If the size of the pump beam is larger than the propagation length of electromagnetic waves in the plasmonic structure, the collective mode is localized in the pumped area.
	In this case, the response time slightly increases with the size of the pump beam.
	In the opposite case, the mode is not localized inside the pumped area and the response time grows with the narrowing of the beam.
	We demonstrate that there is a minimum size of the pump beam for which the modulation speed is a maximum, and can achieve $1$ THz.
	The obtained result opens the way to increasing both the modulation speed and the energy efficiency of plasmonic laser devices.

@@

Here ${{\bf{d}}_m}$ is the dipole moment of $m$th atom at the transition frequency and ${{\bf{E}}_j}({{\bf{r}}_m})$  is the electric field `per one quantum' in $j$th Bloch mode at the position $\textbf{r}_m$ of $m$th atom, see Appendix and Section 2 in Suppl.Mat. of \cite{nefedkin2018acsphot}. Every atom of the active medium has its coordinates $\textbf{r}_m$, which determine the Rabi constants of interaction between the atom and the Bloch modes of the plasmonic structure, see Eq.~(\ref{eq4SM}). This allows us to take into account the spatial distribution of the pumping and the population inversion within the active medium. Assuming that the time of the external pump pulse lies in the femtosecond range and is much shorter than all the characteristic times of the plasmonic DFB laser, in our model, the ultrashort pulse pumping is described as the initial condition, $D_m(0)$. For atoms lying inside the pump beam, the value of $D_m(0)$ is proportional to the pumping energy per unit area. The population inversions of the atoms whose coordinates lie outside the pump beam are $-1$ at the initial time. Note that the square of the Rabi constant is inversely proportional of the Bloch mode volume and is proportional of the Purcell factor of the mode (see Appendix).

@@

For a quantitative description of the temporal dynamics of a plasmonic DFB laser under pulse pumping, we consider a laser based on an Au film perforated by a periodic array of nanosized holes and coated with the layer of the active medium, see Fig.~\ref{fig1}. We consider that the thickness of the Au film is $100$ $\text{nm}$, the diameter of the holes is $160$ $\text{nm}$, and the distance between holes is $470$ $\text{nm}$. The surface of the entire periodic plasmonic structure is $190 \times 190$ $\mu \text{m}^2$. We consider that the transition frequency of the active medium, $\omega_{\sigma}$, is close to the frequency of the second Bloch condition in the plasmonic structure, i.e. $k_\text{pl}(\omega_{\sigma}) \approx G$, where $k_\text{pl}$ is a wavevector of the mode in the plasmonic structure and $G$ is a reciprocal lattice vector \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot}. The linewidth of the active medium is about $100 \text{nm}$. Similar plasmonic DFB lasers have been implemented in a number of experiments \cite{BeijnumPRL,TennerJOpt,TennerACSPhot} and their parameters have been well studied \cite{BeijnumPRL,TennerJOpt,TennerACSPhot,nefedkin2018acsphot}.

The EM field distribution of the Bloch modes of the plasmonic structure have been found within the coupled-mode theory \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot} (see also Appendix), which describes the hybridization of modes of the system without holes through scattering on the holes. Scattering amplitudes for these modes have been taken from \cite{TennerJOpt,TennerACSPhot}. The coupled-mode theory with such scattering amplitudes demonstrates a good agreement with the experimental measurement of the band-structure of the plasmonic nanohole array laser \cite{TennerJOpt,TennerACSPhot}.

In numerical simulations, we take into account $468$ Bloch modes which have different distributions of the EM field inside the cell of the plasmonic structure. The step of quantization of the Bloch wavevectors is determined by the size of the periodic plasmonic structure, see \cite{nefedkin2018acsphot} and Suppl.Mat. of \cite{nefedkin2018acsphot}.

@@

We use the fact of the constancy of the EM field distribution to construct a single-mode model. To do this, we assume that there is a single collective mode, the EM field distribution of which coincides with the EM field distribution found in the numerical simulations of the multimode problem. Furthermore, we assume that the pumped active medium interacts only with such a single collective mode. Within these approximations, the system dynamics can be approximately described by the following system of equations \cite{SiegmanLasers}:
\begin{gather} 
\dot n =  - 2{\gamma _a}n + 2{N_{\text{at}}}{\Omega _R}\varphi \label{ZE1}
\\
\dot \varphi = - \left( {{\gamma _a} + {\gamma _\sigma }} \right)\varphi + {\Omega _R}\left( {2nD + D + 1} \right)/2 \label{ZE2}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 4{\Omega _R}\varphi, \label{ZE3}
\end{gather}
where $n$ is the number of photons in the collective mode, $D$ is the mean population inversion of pumped atoms of the active medium, $\varphi$ describes the mean energy flow from the active medium to the collective mode, and $N_{\text{at}}$ is the number of atoms of the active medium lying in the collective mode.
We suppose that the relaxation rate of the EM field in the collective mode is equal to the characteristic relaxation rate of the modes of the plasmonic structure, i.e., $\gamma_a \simeq 10^{12} \text{s}^{-1}$.
The typical relaxation rate of the population inversion of the active medium is $\gamma_D \simeq 10^9 \text{s}^{-1}$ while the phase relaxation rate is $\gamma_{\sigma}\simeq 10^{13} \text{s}^{-1}$.
Thus, the phase relaxation rate is much larger than other relaxation rates, $\gamma_\sigma \gg \gamma_a, \gamma_D$.
For this reason, one can adiabatically eliminate $\varphi$ from Eqs.~(\ref{ZE1})--(\ref{ZE3}) and obtain the conventional rate equations
\begin{gather} 
\dot n =  - 2 {\gamma _a}n + {N_{\text{at}}}\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE1}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 2\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE2}
\end{gather}
As initial condition we take $D(0) = 1$ and $n(0) = 0$, i.e., all atoms are inverted and there are no photons in the mode.
The temporal dynamics of such a system is well-known \cite{SiegmanLasers}; it is shown in Fig.~\ref{fig4}.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}[h]{0.45\linewidth}
		\includegraphics[width=\linewidth]{Fig5a.eps}
		%\caption{}
	\end{subfigure}
	\begin{subfigure}[h]{0.45\linewidth}
		\includegraphics[width=\linewidth]{Fig5b.eps}
		%\caption{}
	\end{subfigure}
	\caption{(a) The dependence of the delay time on the pump beam diameter. The blue solid line is obtained from numerical simulation of Eqs. (\ref{eq1})--(\ref{eq3}), the orange thick line is obtained from Eq. (\ref{tdel}). (b) The dependence of the diameter of the collective mode on the pump beam diameter (blue line). The black dashed line denotes the pump beam diameter.
	}
	\label{fig5}
\end{figure}

The figure \ref{fig5}a shows that the analytical approximation, Eq.~(\ref{tdel}), agrees well with the numerical simulation of the exact model, Eqs. (\ref{eq1})--(\ref{eq3}).
Let us analyze the dependence of the delay time, Eq.~(\ref{tdel}), on the size of the pumped area.
The number of atoms depends quadratically on the diameter $d_{\text{beam}}$ of the pump beam, i.e., $N_{\text{at}} \sim d_{\text{beam}}^2$.
In turn, the square of the Rabi constant is inversely proportional to the square of the diameter $d_{\text{mode}}^2$ of the collective mode, i.e., $\Omega_{R}^{2} \sim 1/d_{\text{mode}}^2$.
So, we have $t_{\text{delay}} \sim d_{\text{mode}}^2\ln \left( d_{\text{beam}}^2 \right ) /d_{\text{beam}}^2$.
The dependence of $d_{\text{mode}}$ on $d_{\text{beam}}$ is shown in Fig.~\ref{fig5}b.

It can be seen that there is a critical value, $d_{\text{beam}}^*$, of the diameter of the pump beam, which divides two different cases.
When $d_{\text{beam}}>d_{\text{beam}}^*$, the electric field of the collective mode is localized inside the pump beam, see Figs.~\ref{fig3}(d)--(f).
As a consequence, the mode diameter is proportional to the beam diameter, i.e., $d_{\text{mode}} \sim d_{\text{beam}}$, see Fig.~\ref{fig5}b.
For this case, from Eq.~(\ref{tdel}) we have $t_{\text{delay}} \sim \ln d_{\text{beam}}$.
This dependence is in agreement with a slow growth of the delay time of the DFB laser, see Fig\ref{fig2}b.
When $d_{\text{beam}}<d_{\text{beam}}^*$, the electric field of the collective mode is localized outside the pump beam, see Figs.~\ref{fig3}(a)--(c), and the mode size is determined by the characteristic propagation length of the EM field in the unpumped region.
As a result, the mode diameter does not depend on the beam diameter, and $d_{\text{mode}} \sim \text{const}$, see Fig.~\ref{fig5}b.
In this case, $t_{\text{delay}} \sim \ln(d_{\text{beam}}^2)/d_{\text{beam}}^2$.
This is in agreement with an increase of the delay time  when the pump beam diameter of the plasmonic DFB laser is reduced, see Fig.~\ref{fig2}b.
Note that the critical value $d_{\text{beam}}^*$ corresponds to the characteristic propagation length of the EM field outside the pumped active medium and may be evaluated as $d_{\text{beam}}^* \sim v_{\text{group}}/\gamma_a$.
For the parameters of the DFB laser we have $d_{\text{beam}}^* \sim 10 \mu \text{m}$, which is in agreement with the full model, Fig.~\ref{fig2}b.
Thus, because the delay time grows when $d_{\text{beam}}>d_{\text{beam}}^*$ and grows in the opposite case as well, we come to the conclusion that there is a critical value, $\sim d_{\text{beam}}^*$, for the diameter of the pump beam for which the response time is a minimum.

@@

To summarize, we have studied the temporal dynamics of a two-dimensional plasmonic DFB laser under ultrashort pulse pumping.
Relying on numerical simulations and analytical calculations, we have shown that the response time of such a laser for experimental parameters from ~\cite{TennerJOpt,TennerACSPhot} can be decreased to one picosecond, which corresponds to a modulation frequency of $1$ THz.
This value is close to the upper limit for plasmonic lasers (the total response time can not be shorter than the time of the third stage, which is determined by the losses in the cavity, and is $ \sim 1$ ps).

It has been shown that the laser's response time can be shortened by optimizing the size of the pump beam, as a result of the consequent reduction of the delay time.
The dependence of the response time on the size of the pump beam is for following reason.
The interaction of the laser modes with the pumped active medium results in the emergence of a single collective mode.
If the size of the pump beam exceeds the propagation length of the EM waves, this collective mode is located in the pumped area.
The response time then grows logarithmically with increasing pump beam diameter.
In contrast, when the mode is outside the pumped region, the response time is approximately inversely proportional to the square of the pump beam diameter, and so increases rapidly as this diameter decreases from its critical value.
It is interesting to note that the delay time, Eq.~(\ref{tdel}), resembles the dependence of the delay time of a superradiant burst \cite{gross1982superradiance, andreev1980collective,nefedkin2017badcavitySR}.
Indeed, it is well known \cite{gross1982superradiance, andreev1980collective} that a system consisting of initially inverted atoms placed in a low-Q cavity, i.e., when $\gamma_a \gg \gamma_D$, exhibits a superradiant burst with the delay time, Eq.~(\ref{tdel}). In the plasmonic cavity, the inequality $\gamma_a \gg \gamma_D$ is always hold true and, for this reason, the superradiance problem and the problem that we investigate here have the same physical grounds. Further study of the relationship between these problems may be important in the context of creating ultrafast optoelectronic devices.

To sum up, there is the diameter of the pump beam, $d_{\text{beam}}^*$, necessary to achieve the short response time. This diameter is about the decay length of the EM field in the plasmonic structure and for the considered plasmonic laser is $\sim 15 \mu$m. The existence of the critical pump beam size gives the opportunity to reduce the energy consumption because further increase of the pump beam size does not shorten the response time.
In turn, if the area of the pump beam is small, there is the opportunity to reduce the size of the device.
Thus, the obtained results pave the way to creating ultrafast optoelectronic devices based on plasmonic lasers with low power consumption which have the potential of on-chip integration.

@@

\section*{Appendix}
To find the electric field ${{\bf{E}}_j}({{\bf{r}}_m})$  'per one quantum', we equate the energy of one quantum to the energy of the EM field in the resonator:
\begin{equation}
\hbar {\omega _j} = \frac{1}{{8\pi }}\int\limits_V {\left( {{{\left. {\frac{{\partial {\mathop{\rm Re}\nolimits} \left( {\varepsilon(\omega, \textbf{r}) \omega } \right)}}{{\partial \omega }}} \right|}_{\omega  = {\omega _j}}}{{\left| {{{\bf{E}}_j}\left( {\bf{r}} \right)} \right|}^2} + {{\left| {{{\bf{H}}_j}\left( {\bf{r}} \right)} \right|}^2}} \right){d^3}{\bf{r}}} \tag{A.1}
\end{equation}
where $\varepsilon (\omega, \textbf{r})$  is the dielectric constant.
